---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slideimp

<!-- badges: start -->
[![R-CMD-check](https://github.com/hhp94/slideimp/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hhp94/slideimp/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{slideimp}` is a lightweight R package for fast K-NN and PCA imputation of missing values in high-dimensional numeric matrices.

**Core functions**

* `slide_imp()`: Sliding window K-NN or PCA imputation for extremely high-dimensional numeric matrices with ordered features (i.e., by genomic position).
* `group_imp()`: Parallelizable group-wise (e.g., by chromosomes or column clusters) K-NN or PCA imputation with optional auxiliary features and group-wise parameters.
  * (Basic): `group_imp()` requires a `data.frame` (or `tibble`) for the `group` argument. This data frame must contains a `feature` column (character strings of feature names) and a `group` column (defining the group).
  * (Advanced): `group_features()`: `group_imp()`'s helper function to create groups based on a mapping data.frame (i.e., Illumina manifests). See [`{slideimp.extra}`](https://github.com/hhp94/slideimp.extra) on GitHub for tools to process common Illumina manifests.
* `tune_imp()`: Parallelizable hyperparameter tuning with repeated cross-validation; works with built-in or custom imputation functions.
* `knn_imp()`: Full-matrix K-NN imputation with multi-core parallelization, [`{mlpack}`](https://mlpack.org/) KD/Ball-Tree nearest neighbor implementation (for data with very low missing rates and extremely high dimensions), and optional subset imputation (ideal for epigenetic clock calculations).
* `pca_imp()`: Optimized version of [`missMDA::imputePCA()`](http://factominer.free.fr/missMDA/PCA.html) for high-dimensional numeric matrices.

## Installation

The stable version of `{slideimp}` can be installed from CRAN using:

```r
install.packages("slideimp")
```

You can install the development version of `{slideimp}` with:

```r
pak::pkg_install("hhp94/slideimp")
```

## Workflow
Let's simulate some DNA methylation (DNAm) microarray data from 2 chromosomes. All `{slideimp}` functions expect the input to be a numeric matrix where variables are stored in the columns.

```{r}
library(slideimp)
# Simulate data from 2 chromosomes. 20 samples, 50 CpGs
set.seed(1234)
sim_obj <- sim_mat(n = 20, p = 50, perc_total_na = 0.3, perc_col_na = 1, n_col_groups = 2)

sim_obj

# Extract the $input attribute
obj <- sim_obj$input
```

We can estimate the prediction accuracy of different methods and tune hyperparameters prior to imputation with `tune_imp()`.

For custom functions (`.f` argument), the `parameters` data.frame must include the columns corresponding to the arguments passed to the custom function. The custom function must accept `obj` as the first argument and return a matrix with the same dimensions as `obj`.

We tune the results using 2 repeats (`rep = 2`) for illustration (increase in actual analyses).

```{r}
# Our parameters grid contains 2 values: k = 5 and 20.
knn_params <- tibble::tibble(k = c(5, 20))
# Parallelization with OpenMP is controlled by `cores` only for knn or slideimp knn with OpenMP only.
tune_knn <- tune_imp(obj, parameters = knn_params, .f = "knn_imp", cores = 2, rep = 2)
compute_metrics(tune_knn)
```

For K-NN without OpenMP, PCA, and custom functions, setup parallelization with `mirai::daemons()`.
```{r, eval = F}
mirai::daemons(2) # 2 Cores
# Note, for PCA and custom functions, cores is controlled by `mirai::daemons()`
# and the `cores` argument is ignored.

# PCA imputation.
pca_params <- tibble::tibble(ncp = c(1, 5))
tune_pca <- tune_imp(obj, parameters = pca_params, .f = "pca_imp", rep = 2)

# The parameters have `mean` and `sd` columns.
custom_params <- tibble::tibble(mean = 1, sd = 0)
# This function imputes data with rnorm values of different `mean` and `sd`.
custom_function <- function(obj, mean, sd) {
  missing <- is.na(obj)
  obj[missing] <- rnorm(sum(missing), mean = mean, sd = sd)
  return(obj)
}
tune_custom <- tune_imp(obj, parameters = custom_params, .f = custom_function, rep = 2)

mirai::daemons(0) # Close daemons
```

Then, preferably perform imputation by group with `group_imp()` if the variables can be meaningfully grouped (e.g., by chromosomes).

- `group_imp()` allows imputation to be performed separately within defined groups (e.g., by chromosome), which significantly reduces runtime and can increase accuracy for both K-NN and PCA imputation. For the basic API, use a data.frame that contain a `feature` column (character strings of feature names) and a `group` column (defining the group). 

PCA-based imputation with `group_imp()` can be parallelized using the `{mirai}` package, similar to how parallelization is done with `tune_imp()`.

```{r, eval = F}
head(sim_obj$col_group)

# We picked k = 5 (use parameters with lowest errors in actual analyses)
knn_group_results <- group_imp(obj, group = sim_obj$col_group, k = 5, cores = 2)

# Similar to `tune_imp`, parallelization is controlled by `mirai::daemons()`
mirai::daemons(2)
knn_group_results <- group_imp(obj, group = sim_obj$col_group, ncp = 3)
mirai::daemons(0)
```

Alternatively, full matrix imputation can be performed using `knn_imp()` or `pca_imp()`.

```{r, eval = FALSE}
full_knn_results <- knn_imp(obj = obj, k = 5)
full_pca_results <- pca_imp(obj = obj, ncp = 5)
```

## Sliding Window Imputation
Sliding window imputation can be performed using `slide_imp()`. **Note:** DNAm WGBS/EM-seq data should be grouped by chromosomes (i.e. run `slide_imp()` separately on each chromosome) before imputation. See the package vignette for more details.

```{r}
# Simulate some data
chr1_beta <- sim_mat(n = 10, p = 2000, perc_total_na = 0.3, perc_col_na = 1, n_col_groups = 1)$input
dim(chr1_beta)
chr1_beta[1:5, 1:5]
```

`slide_imp()` parameters:

* `location`: **required** - a sorted numeric vector of length `ncol(obj)` giving the position of each column (e.g. genomic coordinates in *bp*).
* `window_size`: **required** - width of each sliding window (same unit as `location`).
* `overlap_size`: **optional** - overlap width between consecutive windows (same units as `location`). Must be strictly less than `window_size`.
* `min_window_n`: **required** - minimum number of columns a window must contain to be imputed. Windows smaller than this are dropped. Must be greater than `k` (for KNN) or `ncp` (for PCA).
* `dry_run`: **optional** - return just the calculated windows to examine which windows are included.
* `k`: **required** - *(specifying KNN imputation)* Number of nearest neighbors to use inside each window.
* `ncp`: **required** - *(specifying PCA imputation)* Number of principal components to retain. Use this instead of `k` when performing sliding-window PCA imputation.
* `subset`: **optional** - impute just a subset of features (i.e., just clock CpGs).
* `flank`: **optional** - to build flanking windows of `window_size` around features provided in `subset`.

First, let's perform a dry run to examine the windows that will be imputed by `slide_imp`.

```{r}
location <- seq_len(ncol(chr1_beta)) # 1, 2, ..., 2000 for this simulated chromosome

slide_imp(
  obj = chr1_beta,
  location = location,
  window_size = 50,
  overlap_size = 5,
  min_window_n = 11, # must be > k = 10
  dry_run = TRUE
)
```

Then, we can perform the imputation using the given parameters.
```{r}
slide_imp(
  obj = chr1_beta,
  location = location,
  window_size = 50,
  overlap_size = 5,
  min_window_n = 11, # must be > k = 10
  k = 10,
  cores = 2,
  .progress = FALSE
)
```
