---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SlideKnn

<!-- badges: start -->
<!-- badges: end -->

`{SlideKnn}` is an efficient R package for k-nearest neighbors (k-NN) imputation of missing values in high-dimensional numeric matrices, such as those from intensive longitudinal data or epi-genomics. It introduces a sliding window approach to handle very large feature sets while preserving local structure, making it suitable for data where features are ordered (e.g., by time or distance).

The package builds on proven k-NN algorithm (Bioconductor's [`impute`](https://www.bioconductor.org/packages/release/bioc/html/impute.html) package) but adds enhancements: parallelization for speed, tree-based methods for efficiency, weighted imputation, multiple imputation strategies, and built-in tuning tools. It's designed for matrices with samples in rows and features in columns.

Key features include:

- **Sliding Window k-NN Imputation**: Break large data into overlapping windows for computationally feasible imputation while maintaining local structures (e.g., intensively sampled longitudinal data or epi-genomic data).
- **Full Matrix k-NN Imputation**: Standard k-NN for smaller data, with multi-core parallelization over columns with missing values.
- **Tree-Based k-NN**: Integration with `{mlpack}` for KD-Tree or Ball-Tree methods, accelerating imputation in high dimensions.
- **Subset Imputation**: Only impute a subset of columns to save time. Important for applications such as epi-genetic clocks calculations.
- **Weighted Imputation**: Use inverse-distance weighting for more accurate averages, with tunable penalties.
- **Multiple Imputation**: Support for Predictive Mean Matching (PMM) or bootstrap resampling from nearest neighbors.
- **Fallback Imputation**: Optional post-k-NN mean imputation by column to handle remaining NAs.
- **Parameter Tuning**: Inject artificial NAs to evaluate and tune hyperparameters, with support for custom imputation functions.
- **Big Matrix Support**: Compatible with `{bigmemory}` for file-backed matrices to handle massive data without loading everything into RAM.

## Installation

The stable version of `{SlideKnn}` can be installed from CRAN using:

```r
install.packages("SlideKnn")
```

You can install the development version of `{SlideKnn}` from [GitHub](https://github.com/hhp94/SlideKnn) with:

``` r
install.packages("remotes")
remotes::install_github("hhp94/SlideKnn")
```

## Example

Load the package and use the built-in `khanmiss1` data (see `?khanmiss1` for details). For full matrix k-NN imputation without sliding windows:

```{r example}
library(SlideKnn)

data(khanmiss1)
# Transpose for samples in rows, features in columns
imputed_full <- knn_imp(t(khanmiss1), k = 3, method = "euclidean", cores = 1)
imputed_full

sum(is.na(imputed_full[[1]]))
```

Yield same results as `{impute::impute.knn()}`. Is faster in larger data and almost as fast in smaller data. Scaling well with multiple `cores`.
```{r}
set.seed(1234)
obj_t <- khanmiss1

bench::mark(
  # Single Core
  knn_imp(t(khanmiss1), k = 3, rowmax = 1, method = "euclidean")[[1]],
  # Multiple Cores
  knn_imp(t(khanmiss1), k = 3, rowmax = 1, cores = 4, method = "euclidean")[[1]],
  t(impute::impute.knn(khanmiss1, k = 3, rowmax = 1, maxp = nrow(khanmiss1))$data)
) |>
  dplyr::mutate(expression = c("knn_1", "knn_4", "impute.knn_1")) |>
  dplyr::select(-dplyr::where(is.list))
```

Sliding window k-NN imputation for epi-genetics data with 1000 CpGs and 10 Samples
```{r}
set.seed(1234)
beta_matrix <- t(sim_mat(m = 10, n = 1000, perc_NA = 0.3, perc_col_NA = 1)$input)
beta_matrix[1:10, 1:5]

imputed <- SlideKnn(beta_matrix, n_feat = 500, n_overlap = 10, k = 10)

imputed
sum(is.na(imputed[[1]][, ]))
```

Using tree-based k-NN with weighting and multiple imputation (PMM):
```{r}
imputed_tree <- knn_imp(
  t(khanmiss1),
  k = 5,
  tree = "kd", # KD-Tree via mlpack
  weighted = TRUE, # Inverse-distance weighting
  dist_pow = 2, # Harsher penalty for distant neighbors
  n_imp = 3, # 3 imputations
  n_pmm = 10 # PMM with 10 donors
)
imputed_tree
```

Optional simple mean imputation as a fallback or baseline that enables multi-step imputation strategies
```{r}
# Inject extra NA into simulated data to make k-NN fails for first imputation
set.seed(1234)
obj <- t(sim_mat(n = 1000, m = 100, perc_NA = 0.8, perc_col_NA = 1)$input)
# Step 1: Column-wise imputation. Disable fall back imputation with `post_imp == FALSE`
imputed_by_col <- knn_imp(obj, cores = 4, k = 10, post_imp = FALSE)
# Step 2: Row-wise imputation. Then if values are still missing, impute by rows
imputed_by_row <- knn_imp(t(imputed_by_col[[1]]), cores = 4, k = 10, post_imp = FALSE)
# Step 3: Lastly, impute by column mean
imputed_mean <- mean_impute_col(t(imputed_by_row[[1]]))
sum(is.na(imputed_mean))
```

For very large matrices that don't fit in memory, use `{bigmemory}` to create file-backed objects. `SlideKnn` supports passing a `big.matrix` object or the path to its descriptor file. Always specify `output` for the result (also file-backed). <b>NOTE</b>: see `?restore_dimnames` if outputs dimnames are stripped.

```{r}
library(bigmemory)
# Set this to allow dimnames in big.matrix
options(bigmemory.allow.dimnames = TRUE)

data(khanmiss1)
mat <- t(khanmiss1) # samples rows, features cols

temp_dir <- withr::local_tempdir()

# Create big.matrix with backing and descriptor files
big_mat <- bigmemory::as.big.matrix(
  mat,
  type = "double",
  backingfile = "khan.bin",
  descriptorfile = "khan.desc",
  backingpath = temp_dir
)

# Impute using the big.matrix object (returns list of big.matrix)
imputed_obj <- SlideKnn(
  obj = big_mat,
  n_feat = 100,
  n_overlap = 10,
  k = 10,,
  overwrite = TRUE, # Overwrite any existing results
  output = file.path(temp_dir, "imputed.bin")
)
sum(is.na(imputed_obj[[1]][, ])) # Access the big.matrix result

# Alternatively, impute using the descriptor path
desc_path <- file.path(temp_dir, "khan.desc")
imputed_path <- SlideKnn(
  obj = desc_path,
  n_feat = 100,
  n_overlap = 10,
  k = 10,
  overwrite = TRUE,
  output = file.path(temp_dir, "imputed.bin")
)

imputed_path
```

## Parameter Tuning
Use `tune_imp()` to tune hyperparameters by injecting artificial NAs and evaluating imputation accuracy. It supports built-in methods or custom functions.

```{r}
parameters <- dplyr::tibble(
  k = c(5, 10),
  method = "euclidean",
  weighted = TRUE,
  dist_pow = c(2, 5),
  post_imp = TRUE
)
parameters

# Subset for faster tuning
obj <- t(khanmiss1)

# 3 repeats, each time inject 50 NAs
results <- tune_imp(obj, parameters, rep = 3, .f = "knn_imp", num_na = 100)

# Compute metrics with {yardstick}
library(yardstick)
met_set <- metric_set(mae, rmse, rsq)
results$metrics <- lapply(results$result, function(x) met_set(x, truth = truth, estimate = estimate))
head(
  dplyr::select(
    tidyr::unnest(dplyr::select(results, -result), cols = "metrics"),
    all_of(names(parameters)), contains(".")
  )
)
```

Tuning a custom imputation function:
```{r}
# This custom function impute NAs with rnorm values
custom_imp <- function(obj, mean = 0, sd = 1) {
  na_pos <- is.na(obj)
  obj[na_pos] <- rnorm(sum(na_pos), mean = mean, sd = sd)
  return(obj)
}

parameters_custom <- dplyr::tibble(
  mean = c(0, 1),
  sd = c(1, 2)
)

results_custom <- tune_imp(obj, parameters_custom, .f = custom_imp, rep = 2, num_na = 20)
results_custom$metrics <- lapply(results_custom$result, function(x) met_set(x, truth = truth, estimate = estimate))
head(
  dplyr::select(
    tidyr::unnest(dplyr::select(results_custom, -result), cols = "metrics"),
    dplyr::all_of(names(parameters_custom)), dplyr::contains(".")
  )
)
```

For more details, see the function documentation (e.g., `?SlideKnn`, `?knn_imp`, `?tune_imp`).
