---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SlideKnn

<!-- badges: start -->
[![R-CMD-check](https://github.com/hhp94/SlideKnn/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hhp94/SlideKnn/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{SlideKnn}` is a lightweight R package for k-nearest neighbors (k-NN) imputation of missing values in high-dimensional numeric matrices, such as those from intensively sampled longitudinal data or epigenetics. `knn_imp()` implements a full k-NN imputation. `SlideKnn()` implements a sliding window k-NN imputation for data where features are ordered (e.g., by time or distance). `group_knn_imp()` implements a group-wise imputation for data that can be broken into meaningful groups, like chromosomes or results of column clustering algorithms.

The package builds on the efficient k-NN imputation algorithm (Bioconductor's [`{impute}`](https://www.bioconductor.org/packages/release/bioc/html/impute.html) package) and adds enhancements: parallelization and tree-based methods for speed, weighted imputation, multiple imputation strategies, and built-in tuning tools. `{SlideKnn}` expects matrices with samples in rows and features in columns.

Key features include:

- **Sliding Window k-NN Imputation**: Break large data into overlapping windows for computationally feasible imputation while maintaining local structures (e.g., intensively sampled longitudinal data or epigenetics data).
- **Group k-NN Imputation**: Break large data into groups such as chromosomes or clusters identified by column clustering algorithms.
- **Full Matrix k-NN Imputation**: Standard k-NN for smaller data, with multi-core parallelization over columns with missing values.
- **Tree-Based k-NN**: Integration with [`{mlpack}`](https://www.mlpack.org/) for KD-Tree or Ball-Tree methods, accelerating imputation in high dimensions.
- **Subset Imputation**: Only impute a subset of columns to save time. Important for applications such as epigenetics clocks calculations.
- **Weighted Imputation**: Use inverse-distance weighting for more accurate averages, with tunable penalties.
- **Multiple Imputation**: Support for Predictive Mean Matching (PMM) or bootstrap resampling from nearest neighbors.
- **Fallback Imputation**: Optional post-k-NN mean imputation by column to handle remaining NAs.
- **Parameter Tuning**: Inject artificial NAs to evaluate and tune hyperparameters, with support for custom imputation functions.
- **Big Matrix Support**: Compatible with [`{bigmemory}`](https://doi.org/10.32614/CRAN.package.bigmemory) for file-backed matrices to handle massive data.

## Installation

The stable version of `{SlideKnn}` can be installed from CRAN using:

```r
install.packages("SlideKnn")
```

You can install the development version of `{SlideKnn}` from [GitHub](https://github.com/hhp94/SlideKnn) with:

``` r
install.packages("remotes")
remotes::install_github("hhp94/SlideKnn")
```

## Full k-NN imputation `knn_imp()`
Use the workhorse function of the package, `knn_imp()` to perform full k-NN imputation.

```{r example}
library(SlideKnn)

data(khanmiss1)
# Transpose for samples in rows, features in columns
imputed_full <- knn_imp(t(khanmiss1), k = 3, method = "euclidean", cores = 1)
imputed_full
```

`knn_imp()` yields the same results as `impute::impute.knn()` without post imputation. `knn_imp()` is faster in larger data and is as fast in smaller data given `cores = 1`. `knn_imp()` scales well with multiple `cores`.
```{r}
set.seed(1234)
obj_t <- t(khanmiss1)

bench::mark(
  # Single Core
  knn_imp(obj_t, k = 3, rowmax = 1, method = "euclidean")[[1]],
  # Multiple Cores
  knn_imp(obj_t, k = 3, rowmax = 1, cores = 4, method = "euclidean")[[1]],
  # impute::impute.knn
  t(impute::impute.knn(khanmiss1, k = 3, rowmax = 1, maxp = nrow(khanmiss1))$data),
  # Ensure results are the same between function calls
  check = TRUE
) |>
  dplyr::mutate(expression = c("knn_1", "knn_4", "impute.knn_1")) |>
  dplyr::select(-dplyr::where(is.list))
```

Using tree-based k-NN with weighted imputation and multiple imputation (PMM):
```{r}
imputed_tree <- knn_imp(
  t(khanmiss1),
  k = 5,
  tree = "kd", # KD-Tree via mlpack
  weighted = TRUE, # Inverse-distance weighting, further neighbors are less influential
  dist_pow = 2, # Harsher penalty for distant neighbors
  n_imp = 3, # 3 imputations
  n_pmm = 10 # PMM with 10 donors. Set n_pmm = 0 to use neighbor bootstrapping instead of PMM
)
imputed_tree
```

## Sliding window k-NN imputation with `SlideKnn()`
Epigenetic datasets such as WGBS or EM-seq are very spatially correlated, this method allows the imputation of the full epigenome with good accuracy and speed by limiting the neighbor search space to sliding windows.
```{r}
# Simulating WGBS data with 1000 CpGs and 10 samples
set.seed(1234)
beta_matrix <- t(sim_mat(m = 10, n = 1000, perc_NA = 0.3, perc_col_NA = 1)$input)
beta_matrix[1:10, 1:5]

imputed <- SlideKnn(beta_matrix, n_feat = 500, n_overlap = 10, k = 10)
imputed
```

Optional simple mean imputation as a fallback or baseline that enables multi-step imputation strategies
```{r}
# Inject extra NA into simulated data to make k-NN fail for first imputation
set.seed(1234)
obj <- t(sim_mat(n = 1000, m = 100, perc_NA = 0.8, perc_col_NA = 1)$input)
# Step 1: Column-wise imputation (impute the features using neighbor features
# of the same person/sample). Disable fall back imputation with `post_imp = FALSE`
imputed_by_col <- knn_imp(obj, cores = 4, k = 10, post_imp = FALSE)
# Step 2: Then if values are still missing, impute by rows. (impute the same features
# using values from OTHER people/samples).
imputed_by_row <- knn_imp(t(imputed_by_col[[1]]), cores = 4, k = 10, post_imp = FALSE)
# Step 3: Lastly, impute by column mean for any remaining missing.
imputed_mean <- mean_impute_col(t(imputed_by_row[[1]]))
sum(is.na(imputed_mean))
```

## Grouped imputation with `group_knn_imp()`
Features can be grouped by chromosomes (in epigenetics) or by cluster memberships identified through column clustering algorithms, such as k-means. This reduces the nearest neighbor search space for each group, significantly decreasing imputation time. The `group` data.frame is defined by two list-columns: `features` and `aux` where each row is a group. The `features` column includes all features to be imputed, while the `aux` columns contains features used to assist with imputation but excluded from the final results. For example, when imputing chrM, which may have only a few CpGs, random CpGs from other chromosomes can be included in the `aux` column to improve the quality of imputation.

```{r}
# Simulate data from 2 chromosomes
set.seed(1234)
to_test <- sim_mat(m = 20, n = 50, perc_NA = 0.3, perc_col_NA = 1, nchr = 2)

# `group_1` will be all the CpGs on Chr1. Same for `group_2`
group_1 <- subset(to_test$group_feature, group == "chr1")$feature_id
group_2 <- subset(to_test$group_feature, group == "chr2")$feature_id

# Impute only first 3 values of group 1, the rest are aux. Group 2 does 4 features.
# Also optionally vary the parameters by group
group_df <- tibble::tibble(
  features = list(group_1[1:3], group_2[1:4]),
  aux = list(group_1, group_2),
  parameters = list(list(k = 3, weighted = TRUE), list(k = 4, method = "manhattan"))
)

group_df

# Run grouped imputation. t() to put features on the columns
obj <- t(to_test$input)
grouped_results <- group_knn_imp(obj, group = group_df, k = 5)
grouped_results
```

## File-backed `big.matrix`
`SlideKnn()` supports passing a `big.matrix` object or the path to its descriptor file. Specify `output` for the result to change the backend to using file-backed big.matrix as well to minimize memory at a cost of performance. 

**NOTE**: See `?restore_dimnames` if output's dimnames are stripped.

```{r}
library(bigmemory)

# IMPORTANT: Enable dimnames support for big.matrix objects
options(bigmemory.allow.dimnames = TRUE)

# Load example data
data(khanmiss1)
mat <- t(khanmiss1) # samples as rows, features as columns
temp_dir <- withr::local_tempdir()
```

### Create file-backed big.matrix
```{r}
# Convert t(khanmiss1) to big.matrix with backing files for large data
big_mat <- bigmemory::as.big.matrix(
  mat,
  type = "double",
  backingfile = "khan.bin",
  descriptorfile = "khan.desc",
  backingpath = temp_dir
)
```

### Method 1: Impute using big.matrix object
Impute a `bigmemory::big.matrix` and optionally save results to file-backed big.matrix
```{r}
imputed_obj <- SlideKnn(
  obj = big_mat,
  n_feat = 100,
  n_overlap = 10,
  k = 10,
  overwrite = TRUE, # Overwrite any existing results
  output = file.path(temp_dir, "imputed.bin")
)

# Access the imputed data (returns list of big.matrix objects)
imputed_obj
```

### ⚠️ Regarding objects pointing to the same bigmemory matrices
On Windows, file-backed objects hold file locks. You MUST remove any variables that point to the same files before overwriting them. This only removes the R variables, not the files on disk.

In general, always remove references to big.matrix objects before attempting to overwrite or modify their backing files.
```{r, echo=TRUE, message=FALSE}
rm(imputed_obj)
invisible(gc(verbose = FALSE)) # Force garbage collection to release file handles
```

### Method 2: Impute using descriptor file path
Alternatively, pass descriptor file path instead of object. This is useful for distributed computing or when the object isn't in memory.
```{r}
desc_path <- file.path(temp_dir, "khan.desc")

imputed_path <- SlideKnn(
  obj = desc_path, # Using path instead of object
  n_feat = 100,
  n_overlap = 10,
  k = 10,
  overwrite = TRUE,
  output = file.path(temp_dir, "imputed.bin")
)

imputed_path
```

## Parameters Tuning with `tune_imp()`
Use `tune_imp()` to tune hyperparameters by injecting artificial NAs and evaluating imputation accuracy. This function supports built-in methods or custom functions.

```{r}
# This tibble defines the hyperparameters to tune for. `knn_imp` requires `k`, but other parameters like `dist_pow` and `method` can also be tuned.
parameters <- dplyr::tibble(
  k = c(5, 10),
  method = "euclidean",
  weighted = TRUE,
  dist_pow = c(2, 5),
  post_imp = TRUE
)
parameters

obj_t <- t(khanmiss1)

# 3 repeats, each time inject 100 NAs
results <- tune_imp(obj_t, parameters, rep = 3, .f = "knn_imp", num_na = 100)

# Compute metrics with {yardstick}
library(yardstick)
met_set <- metric_set(mae, rmse, rsq)
results$metrics <- lapply(results$result, function(x) met_set(x, truth = truth, estimate = estimate))
head(
  dplyr::select(
    tidyr::unnest(dplyr::select(results, -result), cols = "metrics"),
    all_of(names(parameters)), contains(".")
  )
)
```

To tune a custom imputation function, a function has to take an numeric matrix `obj` as the first argument, return a matrix with the same dimension as `obj`, and parameters as arguments.
```{r}
# This custom function imputes NAs with rnorm values. First argument has to be `obj`, which is a numeric matrix. The returned object has to be the same dimension and data type as `obj`.
custom_imp <- function(obj, mean = 0, sd = 1) {
  na_pos <- is.na(obj)
  obj[na_pos] <- rnorm(sum(na_pos), mean = mean, sd = sd)
  return(obj)
}

# This tibble defines the hyperparameters grid similar to the built-in case.
parameters_custom <- dplyr::tibble(
  mean = c(0, 1),
  sd = c(1, 2)
)

# Similarly, use `{yardstick}` to get the prediction metrics
results_custom <- tune_imp(obj, parameters_custom, .f = custom_imp, rep = 2, num_na = 20)
results_custom$metrics <- lapply(
  results_custom$result,
  function(x) met_set(x, truth = truth, estimate = estimate)
)

head(
  dplyr::select(
    tidyr::unnest(dplyr::select(results_custom, -result), cols = "metrics"),
    dplyr::all_of(names(parameters_custom)), dplyr::contains(".")
  )
)
```

For more details, see the function documentation (e.g., `?SlideKnn`, `?group_knn_imp`, `?knn_imp`, `?tune_imp`).

# Developer notes for [`{mlpack}`](https://www.mlpack.org/)

`{mlpack}` is an awesome library and works seamlessly with integration in R. For other developers who are looking to integrate `{mlpack}` into their own work, make sure to structure the .cpp file as follows. Most importantly is the inclusion of `<mlpack.h>` first before including any other `mlpack` headers. This appropriately directs the output of `std::cout` and `std::cerr` to R and won't trip `R CMD check` (i.e., `'_ZSt4cerr', possibly from 'std::cerr' (C++)` or `'_ZSt4cout', possibly from 'std::out' (C++)`). Full implementation is found under `src/impute_knn_mlpack.cpp`. I would like to thank the developers and the community for the maintenance and development of `mlpack`. I would like to especially thank Dirk Eddelbuettel for his development of the R + C++ environment and guidance throughout the development of this package.

```cpp
// [[Rcpp::depends(mlpack, RcppArmadillo)]]
#include <mlpack.h>
#include "imputed_value.h"
#include <mlpack/methods/neighbor_search/neighbor_search.hpp>
#include <RcppArmadillo.h>
```
