---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# slideimp

<!-- badges: start -->
[![R-CMD-check](https://github.com/hhp94/slideimp/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/hhp94/slideimp/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`{slideimp}` is a lightweight R package for fast K-NN and PCA imputation of missing values in high-dimensional numeric matrices.

## Installation

The stable version of `{slideimp}` can be installed from CRAN using:

```r
install.packages("slideimp")
```

You can install the development version of `{slideimp}` with:

```r
pak::pkg_install("hhp94/slideimp")
```

You can install the optional [`{slideimp.extra}`](https://github.com/hhp94/slideimp.extra) package (which provides lightweight Illumina manifests) with:

```r
pak::pkg_install("hhp94/slideimp.extra")
```

## PCA or KNN imputation of Illumina DNAm microarrays
```{r, include = FALSE, echo = FALSE, eval = F}
set.seed(1234)
library(slideimp)
library(slideimp.extra)

set_slideimp_path("dev")
MSA <- ilmn_manifest("MSA")
MSA <- MSA[group != "M"]
n_samples <- 20
MSA_beta_matrix <- slideimp::sim_mat(
  n = n_samples,
  p = nrow(MSA),
  rho = 0.75,
  beta = TRUE
)$input
length_sim <- length(MSA_beta_matrix)
colnames(MSA_beta_matrix) <- MSA$feature
rownames(MSA_beta_matrix) <- paste0("sample", 1:n_samples)
MSA_beta_matrix[1, 1] <- NA
MSA_beta_matrix[3, 2] <- NA
MSA_beta_matrix[1, 4] <- NA
```

* This is a simulated beta matrix from the [MSA](https://www.illumina.com/products/by-type/microarray-kits/infinium-methylation-screening-array.html) microarray. CpGs are in columns.
```{r, eval = F}
dim(MSA_beta_matrix)
# [1]     20 281797
MSA_beta_matrix[1:4, 1:4]
#         cg06185909_TC11 cg18975462_BC11 cg20516119_TC11 cg10149399_BC11
# sample1              NA       0.5023435       0.3835431              NA
# sample2       0.4907466       0.5095459       0.9025816       0.4313347
# sample3       0.6885036              NA       0.7646753       0.4498772
# sample4       0.0000000       0.0000000              NA       0.0000000
```

* Chromosome-wise imputation of Illumina microarrays can be performed with a single function call (requires the Illumina manifests, conveniently provided by the [`slideimp.extra`](https://github.com/hhp94/slideimp.extra) package).

* This example demonstrates PCA imputation. To use KNN imputation instead, supply the `k` argument.

```{r, eval = F}
library(slideimp)
library(slideimp.extra)
library(mirai)

imputed <- group_imp(
  obj = MSA_beta_matrix,
  group = "MSA", # <- this feature requires the `slideimp.extra` package
  ncp = 10, # <- change to `k` for KNN imputation
  .progress = FALSE # <- turn on to monitor progress of longer running jobs
)
# Found cleaned manifest for 'MSA'
# Groups 24 dropped: no features remaining after matching obj columns.
# Imputing 25 group(s) using PCA.
# Running Mode: sequential ...

print(imputed, n = 4, p = 4)
# slideimp_results (PCA)
# Dimensions: 20 x 281797
# 
#         cg06185909_TC11 cg18975462_BC11 cg20516119_TC11 cg10149399_BC11
# sample1       0.1517542       0.5023435      0.38354308       0.2067731
# sample2       0.4907466       0.5095459      0.90258164       0.4313347
# sample3       0.6885036       0.7339375      0.76467530       0.4498772
# sample4       0.0000000       0.0000000      0.05230101       0.0000000
# 
# # Showing [1:4, 1:4] of full matrix
```

* Tips
  - Remove the "ctl" probes (control probes) before imputation with `obj <- obj[, !grepl("^ctl", colnames(obj))]` or set `allow_unmapped = TRUE` to ignore them.
  - Deduplicate the MSA and EPICv2 data after imputation with `slideimp.extra::dedup_matrix()`.
  - Speed up KNN imputation with OpenMP by using the `cores` argument instead of `{mirai}` (available by default on Windows and Linux). If you only need clock CpGs, provide the `subset` argument to ignore all other probes.

## Extended Workflow
* We simulate DNA methylation (DNAm) microarray data from 2 chromosomes. All `{slideimp}` functions expect the input to be a numeric matrix where variables are stored in the columns.

```{r}
library(slideimp)
set.seed(1234)
sim_obj <- sim_mat(n = 20, p = 100, n_col_groups = 2)

# Extract the $input attribute
obj <- sim_obj$input

# Extract metadata to map the features to groups
group_df <- sim_obj$col_group
```

* Hyperparameters are tuned using `tune_imp()`. We evaluate the following options with grid search:

  - Number of nearest neighbors (`k`): 5 or 20
  - Inverse distance power (`dist_pow`) for weighted averaging: 1 or 2
(higher values assign lower weights to more distant neighbors)

* Tuning is performed on a subset of the data. We use 10 repeats (`n_reps = 10`) of cross-validation for evaluation. We re-impute 50 observed values (`num_na = 50`) to compute the rmse and mae. Increase both `n_reps` and `num_na` in real analyses to increase reliability of error estimates.
* **Note:** Parallelization via the `cores` argument is only available for KNN imputation with OpenMP.

```{r}
knn_params <- expand.grid(k = c(5, 20), dist_pow = c(1, 2))
group2_columns <- subset(group_df, group == "group2")
group2_only <- obj[, group2_columns$feature]

tune_knn <- tune_imp(
  group2_only,
  parameters = knn_params,
  .f = "knn_imp",
  cores = 8,
  n_reps = 10,
  num_na = 50
)
```

* Calculate errors using `compute_metrics()` or [`{yardsticks}`](https://github.com/tidymodels/yardstick) functions.
```{r}
metrics <- compute_metrics(tune_knn)

# equivalently: dplyr::summarize(metrics, ..., .by = c(k, dist_pow, .metric))
sum_metrics <- do.call(
  data.frame,
  aggregate(
    .estimate ~ k + dist_pow + .metric,
    data = metrics,
    FUN = function(x) {
      c(
        n = length(x),
        mean_error = mean(x),
        sd_error = sd(x)
      )
    }
  )
)

sum_metrics[order(sum_metrics$.estimate.mean_error), ]
```

* For K-NN without OpenMP, PCA imputation, and custom functions (see vignette), set up parallelization with `mirai::daemons()`.
* **Note:** For machines with multi-threaded BLAS, turn on `pin_blas = TRUE` when tuning PCA imputation in parallel to avoid thrashing.

```{r, eval = F}
mirai::daemons(2) # 2 Cores

# PCA imputation.
pca_params <- data.frame(ncp = c(1, 5))
# For machines multi-threaded BLAS, turn on `pin_blas = TRUE`
tune_pca <- tune_imp(obj, parameters = pca_params, .f = "pca_imp", n_reps = 10, num_na = 50)

mirai::daemons(0) # Close daemons
```

* Then perform imputation with `group_imp()` using the best parameters.
```{r}
knn_group_results <- group_imp(obj, group = group_df, k = 20, dist_pow = 1, cores = 2)
knn_group_results
```

* PCA imputation with `group_imp()` can be parallelized using the `{mirai}` package, similar to `tune_imp()`. `pin_blas = TRUE` can help PCA imputation on multi-threaded BLAS machines.

```{r, eval = F}
# Similar to `tune_imp`, parallelization is controlled by `mirai::daemons()`
mirai::daemons(2)
pca_group_results <- group_imp(obj, group = group_df, ncp = 10)
mirai::daemons(0)
```

* Alternatively, full matrix imputation can be performed using `knn_imp()` or `pca_imp()`.

```{r, eval = FALSE}
full_knn_results <- knn_imp(obj = obj, k = 20)
full_pca_results <- pca_imp(obj = obj, ncp = 10)
```

## Sliding Window Imputation
* `slide_imp()` performs sliding window imputation. 
* **Note:** DNAm WGBS/EM-seq data should be grouped by chromosomes (i.e., run `slide_imp()` separately on each chromosome) before imputation. See the package vignette for more details. 
* This function is **not** for Illumina DNAm microarrays; see `group_imp()`.
* See vignette for details about selecting `window_size` and `overlap_size` with `tune_imp()`.

```{r}
# Simulate some data
chr1_beta <- sim_mat(n = 10, p = 2000)$input
dim(chr1_beta)
chr1_beta[1:5, 1:5]
```

* `slide_imp()` parameters:

  - `location`: **required** - a sorted numeric vector of length `ncol(obj)` giving the position of each column (e.g., genomic coordinates in *bp*).
  - `window_size`: **required** - width of each sliding window (same unit as `location`).
  - `overlap_size`: **optional** - overlap width between consecutive windows (same units as `location`). Must be strictly less than `window_size`.
  - `min_window_n`: **required** - minimum number of columns a window must contain to be imputed. Windows smaller than this are dropped. Must be greater than `k` (for KNN) or `ncp` (for PCA).
  - `dry_run`: **optional** - return just the calculated windows to examine which windows are included.
  - `k`: **required** - *(specifying KNN imputation)* number of nearest neighbors to use inside each window.
  - `ncp`: **required** - *(specifying PCA imputation)* number of principal components to retain. Use this instead of `k` when performing sliding-window PCA imputation.
  - `subset`: **optional** - impute just a subset of features (i.e., just clock CpGs).
  - `flank`: **optional** - build flanking windows of `window_size` around features provided in `subset`.

* First, let's perform a dry run to examine the windows that will be imputed by `slide_imp`.

```{r}
location <- seq_len(ncol(chr1_beta)) # 1, 2, ..., 2000 for this simulated chromosome

slide_imp(
  obj = chr1_beta,
  location = location,
  window_size = 50, # select with `tune_imp()`
  overlap_size = 5,
  min_window_n = 11, # must be > k = 10
  dry_run = TRUE
)
```

* Then, we can perform the imputation using the given parameters.
```{r}
slide_imp(
  obj = chr1_beta,
  location = location,
  window_size = 50, # select with `tune_imp()`
  overlap_size = 5,
  min_window_n = 11, # must be > k = 10
  k = 10, # select with `tune_imp()`
  cores = 2,
  .progress = FALSE
)
```

## Core functions
* `group_imp()`: Parallelizable group-wise (e.g., by chromosomes or column clusters) K-NN or PCA imputation with optional auxiliary features and group-wise parameters. The `group` argument supports:
  * Choices such as `"EPICv2"` or `"MSA"` (requires the  [`{slideimp.extra}`](https://github.com/hhp94/slideimp.extra) package).
  * (Basic): a `data.frame` containing a `feature` column (character strings of feature names) and a `group` column (defining the group).
  * (Advanced): use `prep_groups()` to create groups based on a mapping data.frame (i.e., Illumina manifests) with custom group-wise parameters.
* `slide_imp()`: Sliding window K-NN or PCA imputation for extremely high-dimensional numeric matrices (WGBS/EM-seq) with ordered features (i.e., by genomic position).
* `tune_imp()`: Parallelizable hyperparameter tuning with repeated cross-validation; works with built-in or custom imputation functions.
* `knn_imp()`: Full-matrix K-NN imputation with multi-core parallelization, [`{mlpack}`](https://mlpack.org/) KD/Ball-Tree nearest neighbor implementation (for data with very low missing rates and extremely high dimensions), and optional subset imputation (ideal for epigenetic clock calculations).
* `pca_imp()`: Optimized version of [`missMDA::imputePCA()`](http://factominer.free.fr/missMDA/PCA.html) for high-dimensional numeric matrices.
