---
title: "slideimp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{slideimp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slideimp)
set.seed(1234)
```

## Comparing Methods Using a Shared Cross-Validation Split (`tune_imp()`)
* Let's use the included `khanmiss1` dataset.
```{r}
data(khanmiss1)
obj <- t(khanmiss1)
obj[1:3, 1:3]
```

* Instead of random NA sampling inside `tune_imp()`, which internally calls `sample_na_loc()`, we can generate the NA locations up front and pass them to `tune_imp()`. 
* Here, we randomly select 5 samples from each of 200 random genes for 5 repetitions using `sample_na_loc_stratified()`. To specify just certain columns (i.e., clock CpGs), provide the `subset` argument.
  - `na_loc` have 5 elements (5 repeats) where each row (row and col index) is the position of a missing value.
```{r}
na_loc <- sample_na_loc_stratified(obj, n_cols = 200, n_rows = 5, n_reps = 5)
na_loc[1:3] |> lapply(head)
```

* Then we can compare 1) PCA, 2) KNN imputation, and 3) a custom method, since the cross-validation missing values are the same for all methods.
* **Note**: The custom function requires the first argument to be `obj`, must return an object of the same dimensions, and all subsequent arguments must match the column names of the `parameters` data.frame.
```{r}
# This custom function imputes missing values with random normal values and takes mean and sd as params
rnorm_imp <- function(obj, mean, sd) {
  na <- is.na(obj) 
  obj[na] <- rnorm(sum(na), mean = mean, sd = sd) # <- impute values with rnorm values
  return(obj) # <- return an imputed object with the same dim as obj
}

pca_tune <- tune_imp(
  obj,
  .f = "pca_imp",
  na_loc = na_loc,
  parameters = data.frame(ncp = 10)
)

knn_tune <- tune_imp(
  obj,
  .f = "knn_imp",
  na_loc = na_loc,
  parameters = data.frame(k = 10)
)

rnorm_tune <- tune_imp(
  obj,
  .f = rnorm_imp,
  na_loc = na_loc,
  parameters = data.frame(mean = 0, sd = 1) # must match with arguments of `rnorm_imp`
)
```

* PCA imputation performed best here.
```{r}
mean(compute_metrics(pca_tune)$.estimate)
mean(compute_metrics(knn_tune)$.estimate)
mean(compute_metrics(rnorm_tune)$.estimate)
```

## Group-Wise Imputation with Small-Group Padding and Group-Wise Parameters (`group_imp()`)
* `group_imp()` allows imputation to be performed separately within defined groups (e.g., by chromosome), which significantly reduces runtime and can increase accuracy for both K-NN and PCA imputation.
* `group_imp()` requires the `group` argument, which maps `colnames(obj)` to groups. This can be created up front with `prep_groups()` for advanced features such as group-wise parameters and padding of small groups with random features from other groups. `prep_groups()` returns a list-column data.frame with:
  - `features`: **required** - a list-column where each element is a character vector of variable names to be imputed together.
  - `aux`: **optional** - auxiliary variables to include in each group. These are only used to augment the imputation quality of features and are not imputed themselves. If one group is too small (e.g., chrM), `aux` is used to pad the group by randomly drawing samples from other groups to meet `min_group_size`.
  - `parameters`: **optional** - group-specific imputation parameters.

* First we simulate data from 2 groups. We then create `group3` with only 1 feature to show how `min_group_size` pads it using the `aux` list column.
```{r}
sim_obj <- sim_mat(n = 20, p = 50, n_col_groups = 2)

# Matrix to be imputed
obj <- sim_obj$input
obj[1:6, 1:6]

# Metadata, i.e., which features belong to which group
meta <- sim_obj$col_group
meta[1:5, ]

# We put feature 1 in `group3`
meta[1, 2] <- "group3"
meta[1:5, ]
```

* Then we generate the `group` parameter for `group_imp()` up front. We can see that `group3` has been padded to have 10 columns.
* For each group, we also assign a different number of nearest neighbors as a demonstration.
```{r}
set.seed(1234)
group_imp_df <- prep_groups(colnames(obj), group = meta, min_group_size = 10)
group_imp_df$parameters <- list(list(k = 3), list(k = 4), list(k = 5))
group_imp_df
```

* Finally, we impute `obj` using the modified `group_imp_df`. The `k = 10` passed to `group_imp()` is ignored since all groups have group-wise `k` specified.

```{r}
knn_results <- group_imp(obj, group = group_imp_df, cores = 4, k = 10)
knn_results
```

## Sliding Window Imputation for WGBS/EM-seq Data
### Select `window_size`, `overlap_size`, and PCA/KNN Parameters
* We simulate the output of the `{methylKit}` package. 
  - **Note:** WGBS/EM-seq data should be grouped by chromosome before performing sliding window imputation.
  - Here, we simulate 1000 sites.
* Importantly, we simulate the location vector of each feature (genomic position) with varying spacing of 50 to 500 bp apart from each other.
* The `locations` vector contains the genomic position of each feature (column). It is used to determine which columns are grouped together given a window size.

```{r}
set.seed(1234)
sample_names <- paste0("S", 1:10)
n_sites <- 1000

# Simulate positions with 50–500 bp between each site
distances_between <- sample(50:500, size = n_sites, replace = TRUE)
locations <- cumsum(distances_between) # <- important, location vector

methyl <- data.frame(
  chr = "chr1",
  start = locations,
  end = locations,
  strand = "+"
)

for (i in seq_along(sample_names)) {
  methyl[[paste0("numCs", i)]] <- sample.int(100, size = n_sites, replace = TRUE)
  methyl[[paste0("numTs", i)]] <- sample.int(100, size = n_sites, replace = TRUE)
  methyl[[paste0("coverage", i)]]  <- methyl[[paste0("numCs", i)]] + methyl[[paste0("numTs", i)]]
}

methyl[1:5, 1:10]
```

* First, we convert the data into a beta matrix with features in the columns.
```{r}
numCs_matrix <- as.matrix(methyl[, paste0("numCs", seq_along(sample_names))])
cov_matrix <- as.matrix(methyl[, paste0("coverage", seq_along(sample_names))])
beta_matrix <- numCs_matrix / cov_matrix

colnames(beta_matrix) <- sample_names
rownames(beta_matrix) <- methyl$start

beta_matrix <- t(beta_matrix)
# Set 10% of the data to missing
set.seed(1234)
beta_matrix[sample.int(length(beta_matrix), floor(length(beta_matrix) * 0.1))] <- NA
beta_matrix[1:5, 1:5]
```

* Then, in a real dataset, we would tune hyperparameters using `chr22`. Here, as a demonstration, we use the whole data since the size is small.
* Using 2 repetitions of cross-validation (increase to 10-30 in a real analyses). We are selecting between:
  - `ncp` of 2 or 4, indicating that we are performing sliding PCA imputation. Pass `k` for sliding KNN imputation.
  - `window_size` of 5,000 or 10,000 bp.
  - `overlap_size` fixed at 1,000 bp (does not affect results much in real analyses).

```{r}
params <- expand.grid(ncp = c(2, 4), window_size = c(5000, 10000))
params$overlap_size <- 1000
params$min_window_n <- 20 # windows with less than 20 columns are dropped

# Increase n_reps from 2 in actual analyses and use another chromosome (i.e., chr22)
tune_slide_pca <- tune_imp(
  obj = beta_matrix,
  parameters = params,
  .f = "slide_imp",
  n_reps = 2,
  location = locations
)

metrics <- compute_metrics(tune_slide_pca)

aggregate(.estimate ~ .metric + ncp + window_size, data = metrics, FUN = mean)
```

* Finally, we can use `slide_imp()` to impute the full `beta_matrix`. Use the best parameter combination from the cross-validation metrics.
* First, we use `dry_run = TRUE` to examine the columns to be imputed.
  - `start` and `end` are window location vectors.
  - `window_n` is the number of features included in the column.
```{r}
slide_imp(
  obj = beta_matrix,
  location = locations,
  window_size = 5000,
  overlap_size = 1000,
  ncp = 2,
  min_window_n = 20,
  dry_run = TRUE # <- dry_run to inspect the windows
)
```

* Turn-off `dry_run` to impute the data
```{r}
slide_imp(
  obj = beta_matrix,
  location = locations,
  window_size = 5000,
  overlap_size = 1000,
  ncp = 2,
  min_window_n = 20,
  dry_run = TRUE
)
```

### Subset and Flanking Mode
* Use this mode when you only need to impute specific target features (e.g., clock CpGs) rather than the entire dataset.
  - Pass the desired feature names to the `subset` argument. Only windows containing these features will be imputed.
  - Set `flank = TRUE` to build windows *centered* on each feature in the `subset`. Each window will extend `window_size` bp on either side of the target feature (flanking mode).
  - In this mode, the `overlap_size` argument is ignored.
* Example: Here we only want to impute the features `"1323"` and `"33810"` by creating 5,000 bp flanking windows around each:

```{r}
slide_imp(
  obj = beta_matrix,
  location = locations,
  window_size = 5000,
  ncp = 2,
  min_window_n = 20,
  subset = c("1323", "33810"),
  flank = TRUE,
  dry_run = TRUE
)
```
