---
title: "Multiple imputation of a beta matrix"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slideimp)
```

`pca_imp()` and `knn_imp()` fill each missing value with one best guess. That
is enough for many uses, but an analysis run on a singly imputed matrix treats
the guesses as if they had been measured, so its standard errors come out too
small. Multiple imputation instead draws several plausible values for every
hole, so the uncertainty about the missing cells carries through to the final
estimate.

slideimp does this with a Bayesian low-rank (PCA) model, the method of
Audigier, Husson and Josse (2016) <doi:10.1080/00949655.2015.1104683>. This
article walks through the whole path for a beta-value matrix (DNA methylation,
values in `[0, 1]`):

1. tune the model with `mipca_tune()`,
2. fit it with `mipca_fit()`,
3. check that the chains converged,
4. extract completed matrices with `as.matrix()`, and
5. pool an estimate across them.

## Simulate a beta matrix

The matrix below is built on the logit scale - a mean per CpG with the two
modes typical of methylation, a rank-2 signal, and heavy-tailed noise - and
then mapped into `[0, 1]` with `plogis()`. Half of the columns have 16 of
their 80 samples set to `NA`.

This data follows the model the sampler assumes, so the fit will look good.
It is here to show the steps, not to show how well the method works on real
data.

```{r}
set.seed(2)
n <- 80
p <- 300
mu <- ifelse(runif(p) < 0.5, rnorm(p, -2, 0.7), rnorm(p, 2, 0.7))
U <- matrix(rnorm(n * 2), n, 2)
V <- matrix(rnorm(p * 2, 0, 0.5), p, 2)
noise <- 0.4 * sqrt(4 / 6) * matrix(rt(n * p, df = 6), n, p)
truth <- plogis(matrix(mu, n, p, byrow = TRUE) + U %*% t(V) + noise)
dimnames(truth) <- list(sprintf("s%02d", seq_len(n)), sprintf("cpg%03d", seq_len(p)))

obj <- truth
na_cols <- sample.int(p, p / 2)
for (j in na_cols) obj[sample.int(n, 16), j] <- NA
obj[1:4, 1:4]
mean(is.na(obj))
```

Row and column names are required: the fit uses them to keep track of which
cell is which.

## Tune the model with `mipca_tune()`

Two choices shape the model: the `rank` (how many components) and `nu`, the
degrees of freedom of the noise (`Inf` is Gaussian noise; smaller values allow
heavier tails).

`mipca_tune()` hides some observed cells, imputes them at each candidate rank,
and scores the draws against the hidden truth with the continuous ranked
probability score (CRPS, lower is better). From the same draws it also
estimates the `nu` implied at each rank, so one call settles both.

`transform = "logit_biwhiten"` is the transform for beta values. It maps the
data onto the logit scale (after squeezing exact 0 and 1 inwards so the logit
stays finite) and then rescales rows and columns. The sampler works on that
scale, and every value it hands back is mapped back into `[0, 1]`. The
argument has no default: pass `transform = "identity"` for data that is
already roughly Gaussian.

```{r}
tuned <- mipca_tune(
  obj,
  parameters = data.frame(rank = 1:4),
  transform = "logit_biwhiten",
  seed = 1
)
tuned
```

How to read the grid:

- `crps`: the mean score on the hidden cells. The best rank is where it
  bottoms out.
- `ess10` and `essOK`: effective sample size (the 10th percentile across
  cells, and the share of cells above the ESS bar). Low values mean the chains
  are short, not that the rank is wrong. `update(tuned, iter_sampling = ...)`
  extends every arm.
- `rhatMx`: the worst R-hat across the diagnostic cells. Values well above
  1.05 mean the chains disagree.
- `impl.nu` and `band`: the noise degrees of freedom implied by the draws at
  that rank, with an uncertainty band.

The per-cell scores are in `tuned$cells` if you want to look closer:

```{r}
tapply(tuned$cells$crps, tuned$cells$rank, mean)
```

## Fit the model with `mipca_fit()`

`mipca_fit(tuned)` runs the production fit at the rank with the lowest CRPS,
with `nu` taken from that rank's estimate and the chain settings carried over
from the tuner. A `nu` below 5 is raised to 5, because noise that heavy can
occasionally produce an extreme draw that destabilizes the sampler. The
transform is refitted on the full matrix.

```{r}
fit <- mipca_fit(tuned, seed = 2)
fit
```

By default this runs 4 chains with 100 warm-up and 100 sampling iterations
each, so 400 draws are kept for every missing cell.

To set the rank and `nu` yourself, call `mipca()` directly. `mipca()` also
takes `subset`, which limits which columns have their draws stored. Every
hole is still imputed at every iteration, because the low-rank structure ties
the columns together, so `subset` saves memory, not time. On a genome-scale
matrix that matters: the draws take cells x draws of memory.

## Check convergence

`summary()` reports R-hat and effective sample size for a sample of the
imputed cells.

```{r}
summary(fit)
```

The five numbers for each row are the minimum, lower hinge, median, upper hinge
and maximum across cells. Most cells here have R-hat near 1 and an ESS in the
hundreds, so the chains agree and the 400 draws are worth a few hundred
independent ones. The worst cells are listed below the summary. If many of
them sit above an R-hat of 1.05, or the ESS is low, run longer chains:
`mipca_fit(tuned, iter_sampling = 400)`, for example.

## Extract completed matrices

`as.matrix(fit, draw = b)` returns one completed matrix: the observed values,
with draw `b` filled into the holes. We follow two CpGs with missing values
from here on.

```{r}
two <- names(sort(colSums(is.na(obj)), decreasing = TRUE))[1:2]
completed_1 <- as.matrix(fit, draw = 1)
dim(completed_1)
head(completed_1[, two])
```

Completed matrices are built one at a time, so the draws are never all
expanded into memory at once. A typical analysis takes a few dozen draws
spread across the chains:

```{r}
draws <- round(seq(1, fit$nkeep, length.out = 20))
completed <- lapply(draws, function(b) as.matrix(fit, draw = b)[, two])
length(completed)
```

Each hole now has a distribution rather than a single value. For one missing
cell, here are its draws next to the value that was hidden:

```{r}
hole <- which(is.na(obj[, two[1]]))[1]
cell_draws <- vapply(seq_len(fit$nkeep), function(b) {
  as.matrix(fit, draw = b)[hole, two[1]]
}, numeric(1))
quantile(cell_draws, c(0.05, 0.5, 0.95))
truth[hole, two[1]]
```

## Pool an estimate

Run the analysis on every completed matrix and combine the results with
Rubin's rules. As an example, estimate the mean beta value of the first CpG:

```{r}
est <- vapply(completed, function(m) mean(m[, 1]), numeric(1))
within <- vapply(completed, function(m) var(m[, 1]) / nrow(m), numeric(1))

M <- length(completed)
q_bar <- mean(est)                  # pooled estimate
w <- mean(within)                   # average within-imputation variance
b <- var(est)                       # between-imputation variance
total <- w + (1 + 1 / M) * b        # total variance

c(estimate = q_bar, se = sqrt(total), se_single = sqrt(w))
mean(truth[, two[1]])
```

`se_single` is the standard error you would report from one completed matrix.
The pooled `se` adds the between-imputation variance `b`, which is the
uncertainty about the missing cells that single imputation leaves out. Here
only 16 of 80 values are missing and the model fits well, so the difference
is small; it grows with the share of missing values.

## Reference

Audigier, V., Husson, F. and Josse, J. (2016). Multiple imputation for
continuous variables using a Bayesian principal component analysis. *Journal
of Statistical Computation and Simulation*, 86(11), 2140-2156.
<doi:10.1080/00949655.2015.1104683>
