---
title: "Speeding up PCA imputation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slideimp)
```

# Tips to speed up PCA imputation

`pca_imp()` is the PCA imputation workhorse used by `slide_imp()`, `tune_imp()`, and `group_imp()` whenever PCA imputation is requested. This article describes how to make PCA imputation faster by choosing the convergence threshold, the PCA eigensolver, and a parallel backend with `{mirai}`.

## Overview

Most users can keep the defaults:

```{r}
set.seed(1234)
sim <- sim_mat(
  n = 80,
  p = 1500,
  rho = 0.7,
  perc_total_na = 0.10,
  perc_col_na = 0.8
)
sim_df <- sim$col_group
obj <- sim$input
pca_imp(obj, ncp = 2)
```

However, for large genomic matrices, runtime can depend strongly on two arguments:

* `threshold`: the convergence threshold for the PCA imputation algorithm.
  - The default is `threshold = 1e-6`, matching the conservative behavior of `{missMDA}`.
  - For many genomic matrices, `threshold = 1e-5` can be much faster while producing very similar imputed values.
* `solver`: the eigensolver used inside `pca_imp()`.
  - `solver = "auto"` runs a short timed probe of both solvers and picks the faster one.
  - `solver = "exact"` forces the exact solver.
  - `solver = "lobpcg"` forces the LOBPCG with warm start solver.

## Exact versus LOBPCG with warm start solvers

`pca_imp()` supports two PCA eigensolvers:

* the exact solver, selected by `solver = "exact"`;
* the LOBPCG solver with warm start, selected by `solver = "lobpcg"`.

LOBPCG has two advantages over the exact solver. First, it scales well when only a small number of `ncp` is needed and the matrix is approximately low ranked. Second, it benefits from warm starts: in the iterative EM algorithm, the eigensolver is called repeatedly, and consecutive solutions usually differ only slightly as the imputed values stabilize. `pca_imp()` reuses information from the previous solution to warm start subsequent LOBPCG calls.

As a result, while the exact solver is robust and faster for small to moderate matrices, the LOBPCG solver with warm start can be much faster for large, approximately low-rank matrices such as DNAm beta matrices, especially when only a small number of principal components is needed.

The default `solver = "auto"` runs a short timing probe of the available
solvers and chooses LOBPCG only when it is clearly faster. For most users,
`solver = "auto"` is a good default. But, consider setting the solver manually
when:

* You are running many independent `pca_imp()` calls, i.e., in `tune_imp()` or
  `group_imp()`, where repeated probing may add overhead.
* You are working with matrices where you know one solver should dominate but
  `"auto"` is selecting the wrong solver.

## Tuning workflow

Before an expensive analysis, a practical `pca_imp()` tuning workflow is:

1. Choose a representative subset of the data, for example chromosome 22.
2. Time the `exact` and `lobpcg` solvers on the subset at a fixed `threshold`.
3. Use `tune_imp()` to check that a relaxed `threshold` does not materially change cross-validation error.
4. Use the chosen `solver` and `threshold` when tuning accuracy-related parameters, i.e., `ncp` and `coeff.ridge` (and `window_size` and `overlap_size` for `slide_imp()`).

We demonstrate this workflow as follows.

## Create a representative test matrix

In a real DNA methylation analysis, use a representative chromosome or subset. Here, we treat `group1` of the simulated data as the representative chromosome.

```{r}
set.seed(1234)

obj_probe <- obj[, subset(sim_df, group == "group1")$feature]

dim(obj_probe)
mean(is.na(obj_probe))
```

## Time solver and threshold settings

First, choose a small `ncp` for the speed probe. This does not need to be the final value of `ncp`. The goal is to identify whether the exact or iterative solver is faster on this kind of data. We use a relaxed `threshold = 1e-5` to keep the probe quick. Both solvers use the same convergence criterion, so their relative timing should be representative.

```{r}
ncp_probe <- 5
```

```{r}
exact <- system.time(invisible(
  pca_imp(
    obj = obj_probe,
    ncp = ncp_probe,
    solver = "exact",
    threshold = 1e-5,
    na_check = FALSE
  )
))

iterative <- system.time(invisible(
  pca_imp(
    obj = obj_probe,
    ncp = ncp_probe,
    solver = "lobpcg",
    threshold = 1e-5,
    na_check = FALSE
  )
))

exact - iterative
```

Second, use `tune_imp()` to compare cross-validation error across `threshold` values. In this example, relaxing the threshold to `1e-5` does not meaningfully change the error.

```{r}
thresh_df <- data.frame(threshold = c(1e-5, 1e-6), ncp = ncp_probe)
res <- tune_imp(obj = obj_probe, parameters = thresh_df, .f = "pca_imp")

compute_metrics(res)
```

If `threshold = 1e-5` changes cross-validation error or imputed values more than expected, use the more conservative `threshold = 1e-6`.

If neither solver is clearly faster on the representative subset, pick `solver = "exact"` for robustness. If one solver is clearly faster, force it explicitly to skip the auto probe overhead.

## Use the chosen settings with `tune_imp()`

After choosing `solver` and `threshold`, keep them fixed while tuning accuracy-related PCA parameters (i.e., `ncp` and `coeff.ridge`):

```{r}
chosen_solver <- "exact"
chosen_threshold <- 1e-5

params <- expand.grid(
  ncp = c(2, 4, 6),
  coeff.ridge = c(0.8, 1, 1.2)
)

params$solver <- chosen_solver
params$threshold <- chosen_threshold

tune_pca <- tune_imp(
  obj = obj_probe,
  .f = "pca_imp",
  parameters = params,
  n_reps = 1 # <- increase to a higher number in real analyses
)

metrics <- compute_metrics(tune_pca, metrics = "rmse")

aggregate(
  .estimate ~ ncp + coeff.ridge,
  data = metrics,
  FUN = mean
)
```

## Use the chosen settings with `slide_imp()`

The same idea applies when tuning sliding-window PCA imputation. First choose the speed settings on a representative subset, then include those fixed settings in the `parameters` data frame while tuning window and PCA parameters.

`slide_imp()` typically operates on small windows with few samples, so the `"exact"` solver is usually the better default. If `solver = "auto"`, `slide_imp()` will only probe the first window and force the solver for the rest of the windows.

```{r}
beta_matrix <- obj_probe # Treat `obj_probe` as a `slide_imp()` input
locations <- seq_len(ncol(beta_matrix))
slide_params <- expand.grid(
  ncp = c(2, 4),
  coeff.ridge = c(0.8, 1, 1.2),
  window_size = c(50, 500) # Adjust based on units of `locations`
)

slide_params$overlap_size <- 5
slide_params$min_window_n <- 20
slide_params$solver <- "exact" # Usually better for `slide_imp()`
slide_params$threshold <- chosen_threshold

tune_slide_pca <- tune_imp(
  obj = beta_matrix,
  .f = "slide_imp",
  parameters = slide_params,
  n_reps = 1, # <- increase to a higher number in real analyses
  location = locations
)

slide_metrics <- compute_metrics(tune_slide_pca, metrics = "rmse")

aggregate(
  .estimate ~ ncp + coeff.ridge + window_size,
  data = slide_metrics,
  FUN = mean
)
```

Finally, use the selected speed and accuracy parameters when imputing the full dataset:

```{r}
slide_results <- slide_imp(
  obj = beta_matrix,
  location = locations,
  window_size = 500,
  overlap_size = 5,
  min_window_n = 20,
  ncp = 2,
  coeff.ridge = 1.2,
  solver = "exact", # Usually better for `slide_imp()`
  threshold = chosen_threshold,
  .progress = FALSE
)
```

## Parallel PCA imputation and BLAS threading

PCA imputation can also be sped up by running independent imputation tasks in parallel with `tune_imp()` or `group_imp()`. For PCA imputation, we use `{mirai}` parallelization rather than the `cores` argument.

```{r}
library(mirai)

# Start 4 mirai workers.
mirai::daemons(4)

# Run PCA tuning in parallel.
tune_pca <- tune_imp(
  obj = obj_probe,
  .f = "pca_imp",
  parameters = params,
  n_reps = 1,
  .progress = FALSE # <- Use `TRUE` to monitor longer-running jobs
)

# Stop mirai workers when finished.
mirai::daemons(0)
```

When PCA imputation is run in parallel, each worker may call multithreaded linear algebra routines through BLAS. If each worker uses multiple BLAS threads, the total number of active CPU threads can become much larger than the number of physical cores. This is called **thread oversubscription**, and it can make code slower rather than faster.

For example, 8 `{mirai}` workers each using 8 BLAS threads can try to use 64 CPU threads.

To avoid this, set `pin_blas = TRUE` in `tune_imp()` or `group_imp()` when running PCA imputation with `{mirai}`. This requires `{RhpcBLASctl}`:

```{r}
mirai::daemons(4)

tune_slide_pca <- tune_imp(
  obj = beta_matrix,
  .f = "slide_imp",
  parameters = slide_params,
  n_reps = 1,
  location = locations,
  pin_blas = TRUE # <- Set to `TRUE` if R is linked to multi-threaded BLAS/LAPACK
)

mirai::daemons(0)
```

For group-wise PCA imputation:

```{r}
mirai::daemons(4)

pca_group_results <- group_imp(
  obj = obj,
  group = sim_df,
  ncp = 2,
  solver = chosen_solver,
  threshold = chosen_threshold,
  pin_blas = TRUE # <- Set to `TRUE` if R is linked to multi-threaded BLAS/LAPACK
)

mirai::daemons(0)
```

`pin_blas = TRUE` uses `{RhpcBLASctl}` to limit the number of BLAS threads inside parallel workers. This works with common BLAS libraries such as OpenBLAS and MKL, which are typical on Linux and are also available on Windows after replacing R's default BLAS. It may have no effect on macOS, depending on the BLAS implementation.

### BLAS/LAPACK swapping for advanced Windows users

On Windows machines, R’s default BLAS/LAPACK is reliable but single-threaded and can be slow. Advanced users can get much faster PCA imputation by replacing R’s default BLAS with an optimized BLAS following this guide: [OpenBLAS for R on Windows](https://github.com/david-cortes/R-openblas-in-windows).

However, OpenBLAS is multithreaded, so for parallel runs using `{mirai}`, set `pin_blas = TRUE` to avoid thread oversubscription.

## Practical recommendations

* Start with `solver = "auto"` unless PCA imputation is slow or you are about to run many calls, such as with `tune_imp()`, where the per-call probe overhead can add up.
* If PCA imputation is slow, benchmark `solver = "exact"` against `solver = "lobpcg"` on a representative subset and force the faster solver.
* Try `threshold = 1e-5` or a less stringent value, depending on your error tolerance.
* Keep `threshold = 1e-5` only if it produces similar imputed values or similar cross-validation error.
* LOBPCG is most likely to help when the matrix is large, approximately low-rank, and `ncp` is small.
* For `slide_imp()`, `solver = "exact"` is usually the right choice.
* Choose speed settings first, then tune accuracy-related parameters such as `ncp`, `coeff.ridge`, and `window_size`.
* If running PCA imputation sequentially, BLAS multithreading may help, and `pin_blas` is unnecessary.
* If running many PCA imputations in parallel with `{mirai}`, use `pin_blas = TRUE`.
* Avoid using many `{mirai}` workers and many BLAS threads at the same time.
* Benchmark on a representative subset, because the best setup depends on the hardware, BLAS library, matrix size, and missing-data pattern.

