---
title: "Speeding up PCA imputation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(slideimp)
```

# Tips to speed up PCA imputation

`pca_imp()` is the PCA imputation workhorse used by `slide_imp()`,
`tune_imp()`, and `group_imp()` whenever PCA imputation is requested. This
article describes how to make PCA imputation faster by choosing `threshold`,
`scale`, `solver`, and a parallel backend with `{mirai}`.

## Overview

For large genomic matrices, runtime can be significantly reduced by tuning
three arguments: `threshold`, `scale`, and `solver`.

```{r}
set.seed(1234) # simulate some data
sim <- sim_mat(
  n = 80,
  p = 1500,
  rho = 0.7,
  perc_total_na = 0.10,
  perc_col_na = 0.8
)
sim_df <- sim$col_group
obj <- sim$input
```


### `threshold`

The convergence threshold for the PCA imputation algorithm. The default
`threshold = 1e-6` matches the conservative behavior of `{missMDA}`. For many
genomic matrices, `threshold = 1e-5` is much faster while producing very
similar imputed values.

### `scale`

Whether columns are centered and scaled before PCA. The default `scale = TRUE` 
adds per-iteration overhead. For data where columns are on the same scale
(e.g., DNAm beta values in `[0, 1]`), `scale = TRUE` may amplify low-variance and 
noise-dominated columns. For such data, `scale = FALSE` can be both faster 
and more accurate.

### `solver`

`pca_imp()` supports two PCA eigensolvers:

* `solver = "exact"`: the exact solver. Robust and typically faster for small
  to moderate matrices.
* `solver = "lobpcg"`: LOBPCG with warm start. Scales well when only a small
  `ncp` is needed and the matrix is approximately low rank. In the iterative
  EM algorithm, consecutive eigensolver calls differ only slightly as imputed
  values stabilize. `pca_imp()` **warm-starts LOBPCG with both the previous
  eigenblock and search direction**, so subsequent calls converge in few
  iterations.
* `solver = "auto"` (default): runs a short timed probe of both solvers and
  picks LOBPCG only when it is clearly faster.

For most users, `solver = "auto"` is a good default. But, consider setting the
solver manually when:

* You are running many independent `pca_imp()` calls (i.e., in `tune_imp()`
  or `group_imp()`), where repeated probing adds overhead.
* You know one solver should dominate but `"auto"` is selecting the wrong one.

If neither solver is clearly faster on a representative subset, pick
`solver = "exact"` for robustness.

## Tuning workflow

Before an expensive analysis, a practical `pca_imp()` tuning workflow is:

1. Choose a representative subset of the data, for example chromosome 22.
2. Time the `exact` and `lobpcg` solvers on the subset at a fixed `threshold`.
3. Use `tune_imp()` to check that a relaxed `threshold` and `scale = FALSE` do
   not materially change cross-validation error.
4. Fix `solver`, `threshold`, and `scale`, then tune the other parameters
   (`ncp` and `coeff.ridge`, and `window_size` and `overlap_size` for
   `slide_imp()`).

### Create a representative test matrix

In a real DNA methylation analysis, use a representative chromosome or subset.
Here, we treat `group1` of the simulated data as the representative chromosome.

```{r}
set.seed(1234)

obj_probe <- obj[, subset(sim_df, group == "group1")$feature]

dim(obj_probe)
mean(is.na(obj_probe))
```

### Time the solvers

Use a small `ncp` and a relaxed `threshold = 1e-5` for the probe. This `ncp` is 
not necessary the final `ncp`. Both solvers use the same convergence criterion, 
so their relative timing is representative.

```{r}
ncp_probe <- 5

exact <- system.time(invisible(
  pca_imp(
    obj = obj_probe,
    ncp = ncp_probe,
    solver = "exact",
    threshold = 1e-5,
    na_check = FALSE
  )
))

iterative <- system.time(invisible(
  pca_imp(
    obj = obj_probe,
    ncp = ncp_probe,
    solver = "lobpcg",
    threshold = 1e-5,
    na_check = FALSE
  )
))

# diff
exact[["elapsed"]] - iterative[["elapsed"]]
```

### Check `threshold`

Use `tune_imp()` to compare cross-validation error across `threshold` values.
In this example, relaxing the threshold to `1e-5` does not meaningfully change
the error.

```{r}
thresh_df <- data.frame(threshold = c(1e-5, 1e-6), ncp = ncp_probe)
res <- tune_imp(obj = obj_probe, parameters = thresh_df, .f = "pca_imp")

compute_metrics(res)
```

### Check `scale`

`scale = FALSE` can be both faster and more accurate for certain DNAm data. We
verify this with `tune_imp()`.

```{r}
scale_df <- data.frame(
  scale = c(TRUE, FALSE),
  ncp = ncp_probe,
  threshold = 1e-5
)
res_scale <- tune_imp(obj = obj_probe, parameters = scale_df, .f = "pca_imp")

compute_metrics(res_scale)
```

### Tune accuracy parameters with the chosen settings

After choosing `solver`, `threshold`, and `scale`, keep them fixed while
tuning accuracy-related PCA parameters (`ncp` and `coeff.ridge`). Here, the 
`solver = "lobpcg"` is not clearly faster and `scale = FALSE` is more accurate.

```{r}
chosen_solver <- "exact"
chosen_threshold <- 1e-5
chosen_scale <- FALSE

# tune ncp and coeff.ridge
params <- expand.grid(
  ncp = c(2, 4, 6),
  coeff.ridge = c(0.8, 1, 1.2)
)

params$solver <- chosen_solver
params$threshold <- chosen_threshold
params$scale <- chosen_scale

tune_pca <- tune_imp(
  obj = obj_probe,
  .f = "pca_imp",
  parameters = params,
  n_reps = 1 # <- increase to a higher number in real analyses
)

metrics <- compute_metrics(tune_pca, metrics = "rmse")

aggregate(
  .estimate ~ ncp + coeff.ridge,
  data = metrics,
  FUN = mean
)
```

### Apply to `slide_imp()`

The same idea applies when tuning sliding-window PCA imputation. `slide_imp()`
typically operates on small windows with few samples, so `solver = "exact"` is
usually the right choice. With `solver = "auto"`, `slide_imp()` probes only
the first window and reuses the chosen solver for the rest.

```{r}
beta_matrix <- obj_probe # Treat `obj_probe` as a `slide_imp()` input
locations <- seq_len(ncol(beta_matrix))
slide_params <- expand.grid(
  ncp = c(2, 4),
  coeff.ridge = c(0.8, 1, 1.2),
  window_size = c(50, 500) # Adjust based on units of `locations`
)

slide_params$overlap_size <- 5
slide_params$min_window_n <- 20
slide_params$solver <- "exact" # Usually better for `slide_imp()`
slide_params$threshold <- chosen_threshold
slide_params$scale <- chosen_scale

tune_slide_pca <- tune_imp(
  obj = beta_matrix,
  .f = "slide_imp",
  parameters = slide_params,
  n_reps = 1, # <- increase to a higher number in real analyses
  location = locations
)

slide_metrics <- compute_metrics(tune_slide_pca, metrics = "rmse")

aggregate(
  .estimate ~ ncp + coeff.ridge + window_size,
  data = slide_metrics,
  FUN = mean
)
```

Finally, use the selected parameters when imputing the full dataset:

```{r}
slide_results <- slide_imp(
  obj = beta_matrix,
  location = locations,
  window_size = 500,
  overlap_size = 5,
  min_window_n = 20,
  ncp = 2,
  coeff.ridge = 1.2,
  solver = "exact",
  threshold = chosen_threshold,
  scale = chosen_scale,
  .progress = FALSE
)
```

## Parallel PCA imputation and BLAS threading

PCA imputation can also be sped up by running independent imputation tasks in
parallel with `tune_imp()` or `group_imp()`. For PCA imputation, we use
`{mirai}` parallelization rather than the `cores` argument.

When PCA imputation is run in parallel, each worker may call multithreaded
linear algebra routines through BLAS. If each worker uses multiple BLAS
threads, the total number of active CPU threads can far exceed the number of
physical cores, for example, 8 `{mirai}` workers each using 8 BLAS threads
will request 64 threads. This **thread oversubscription** can make code slower
rather than faster.

Set `pin_blas = TRUE` to avoid this. It uses `{RhpcBLASctl}` to limit BLAS
threads inside workers, and works with common BLAS libraries such as OpenBLAS
and MKL.

```{r}
library(mirai)

mirai::daemons(4)

pca_group_results <- group_imp(
  obj = obj,
  group = sim_df,
  ncp = 2,
  solver = chosen_solver,
  threshold = chosen_threshold,
  scale = chosen_scale,
  pin_blas = TRUE # <- Set to `TRUE` if R is linked to multi-threaded BLAS/LAPACK
)

mirai::daemons(0)
```

The same `pin_blas = TRUE` argument is available in `tune_imp()`.

## Practical recommendations

* Start with `solver = "auto"` unless you are about to run many calls (e.g.,
  with `tune_imp()`), where per-call probe overhead adds up.
* If PCA imputation is slow, benchmark `"exact"` vs `"lobpcg"` on a
  representative subset and force the faster solver. LOBPCG is most likely to
  help when the matrix is large, approximately low-rank, and `ncp` is small.
* Try `threshold = 1e-5` if it produces similar imputed values or cross-validation error.
* For data where columns already share a common scale (e.g., DNAm beta
  values), try `scale = FALSE` and verify with cross-validation.
* For `slide_imp()`, `solver = "exact"` is usually the right choice.
* Choose speed settings first, then tune accuracy parameters such as `ncp`,
  `coeff.ridge`, and `window_size`.
* When running PCA imputation in parallel with `{mirai}` and a multithreaded
  BLAS, use `pin_blas = TRUE` to avoid thread oversubscription.
* Benchmark on a representative subset. The best setup depends on the
  hardware, BLAS library, matrix size, and missing-data pattern.
